QueryMed: An Intuitive SPARQL Query Builder for Biomedical RDF Data

### 2010-02-12 docs/www2010/paper.tex
general edits to paper

In the following sections we explain the system functionality by first giving an overview of the QueryMed system architecture and the design decisions behind some of the crucial components of the system.

@@

The QueryMed user interface is designed to be intuitive for the end user, yet flexible enough to permit a broad range of interesting queries.  The basic query interface allows the user to run simple queries, and is designed for maximal easy of use.  The advanced search capabilities allow the user to easily construct complex logical queries that take advantage of the underlying structure of the biomedical ontologies.  The user interface also allows the user to iteratively refine queries, and displays the query results, which have been retrieved from multiple SPARQL endpoints.  The main components of the user interface are as follows:

@@

\item \textbf{Basic Query Interface:}  In order to make QueryMed easy to use, but still provide a flexible tool that is able to perform a range of queries, we provide an uncluttered interface with only a  blank search box and two buttons as shown in Figure \ref{fig:initial_ui}. The ``Query All" button will perform a keyword-based query over pre-selected data sources. The advanced search option ``Refine Query" allows the user to view additional query options that are handled in the Advanced Query Interface.

@@

For example, after searching for "coronary artery disease" in the basic query interface, the physician will see displayed in the Result Browser a list of disease names in the Diseasome database and drugs in DailyMed and Drugbank that relate to coronary artery 
disease. She can then filter the results using additional search terms. For example, she knows that 
the route of administration of the drug that she is looking for is injection, so she filters the drug query results on the route of administration field using the query term “injection.”   The additional filtering capabilities of the Result Browser allows her  to either browse a large number of query results, or refine her search to view only the most relevant query results.

\item \textbf{Advanced Query Interface:} This allows the user to add new data sources, and give values for properties found at a particular selected data source. Therefore, the user is able to extend the query from a basic keyword search over all the properties as with the basic query interface, to a more targeted query.

When the ``Refine Query" in the Basic Query Interface is invoked, the user is provided with the default list of data sources in the system as shown in Figure \ref{fig:select}.  This feature allows the user to perform their query only over relevant or trusted data sources.  If none of the desired data sources are available in this list, the user can dynamically add additional data sources as shown in Figure \ref{fig:add_source}.  The ability to dynamically add additional data sources increases the flexibility of  the QueryMed system, allowing users to query endpoints of interest that are not included in the default list.

@@

When a data source is selected, the properties list is automatically populated with all the distinct properties available at the selected endpoint. If the user wants to restrict her query based on one or more properties available, she can specify those. This allows users to improve their queries using the underlying structure of the data without having prior knowledge of this structure to obtain more relevant query results. 
%The properties list is generated by running a SPARQL query of the form given in Figure \ref{fig:SPARQL_props} at the specified data source.

@@

The user can choose to perform exact queries, or filter results on specific keywords. If the user does not know the specific value for a property, they can instead specify a keywords using the “FILTER” option. They also can choose logical operators to connect the various parts of their query. When “AND” is used it will be appended as a basic triple pattern to the SPARQL query, i.e. as a conjunction. When “OR” is used, the specified graph pattern is made to disjunct with the rest of the query with the SPARQL UNION operator. If no logical operator is specified, “AND” will be used by default.  The advanced query feature is capable of dynamically constructing complex SPARQL queries, such as the query shown in Figure \ref{fig:SPARQL_adv}.

@@

For example, the physician might be interested in finding relevant clinical trials for 
her patient. However, the clinical trial database LinkedCt is 
not in the default set of endpoints, so she selects the “Refine 
Query” option to choose additional endpoints to search. She 
sees a list of default endpoints, and selects the “Add” option 
to include an additional endpoint. After entering the name 
and URL of the LinkedCT endpoint, she is able to search 
for clinical trials for which her patient may be eligible.

\item \textbf{Dynamic Facet Creator/Filter:} This allows the user to select a  set of  data sources, load the properties available at these sources, and dynamically construct a complex SPARQL query connected by logical operators to take advantage of the structure of the data at each endpoint as shown in Figure  \ref{fig:facet_creator}. The interface allows displaying of many property lists from different sources simultaneously (grouped by source). Therefore, only information relevant to the endpoint the user is currently examining will be visible at any given time.

For example, the physician might want to further refine her search for cardiovascular diseases. She 
uses the advanced search option to construct a boolean query 
that takes advantage of the underlying structure of the RDF 
data in the database. She searches Diseasome for a list of 
diseases whose class is “Cardiovascular” or for which the as- 
sociated gene is “ABCA1.” Using the QueryMed advanced 
search interface, the complex SPARQL query corresponding 
to her question is automatically constructed (Figure \ref{fig:SPARQL_adv}), and she can 
view the query results conveniently displayed in the Results Browser.

@@

The functionality of the proxy server is to act as an intermediary between the user interface and the remote SPARQL endpoints.  Its functionality is twofold:

\begin{enumerate}
\item Execute the SPARQL queries at the relevant remote SPARQL endpoints and consolidate the results to be presented in the user interface.
\item Cache the results for subsequent queries for refinements to the same query to reduce the network and query execution latency.
\end{enumerate}

@@

\item \textbf{Translator:} The Translator is responsible for translating the user query into valid SPARQL syntax.  Translator obtains the parameters to construct the query from the input the user specifies in the Query Interface, and plugs them in to a relevant template. It specifically relies on two services:

@@

\item \textbf{Run Query:} The purpose of this service is to execute SPARQL queries generated by the Translator, and return  the query results as a JSON object given in Figure \ref{fig:JSON_runquery}.

\begin{figure}
\begin{Verbatim}[frame=single]
{
   "bindings":
     [
        {"source" : "source_label",
         "uri" : "http://source.com/sparql.",
     	"vars" : ["variable1","variable2"],
     	"count":2,
     	"results" : 
     	 [
     	    {"variable1": "Value 1",
	      "variable1": "Value 2"},
     	    {"variable1": "Another Value 1",
               "variable1": "Another Value 2"},
   	    ...	  		
     	 ]
     	}
     	,
	{ //Another Source and Result Description
	...
	},
	...
     ]
 }
\end{Verbatim}
\caption{Structure of the JSON ``result" object}
\label{fig:JSON_runquery}
\end{figure}

\item \textbf{Advanced Queries:} This service takes as input list of sources, properties, query terms, and logical operators, which are passed to the proxy server as a JSON object given in Figure \ref{fig:JSON_advquery}.

\begin{figure}
\begin{Verbatim}[frame=single]
{
 "http://some-source/sparql": 
  {
     "property1": {"value1":  "logical operator"},
     "property2": {"value 2": "logical operator"},
     "property2": {"value 3": "logical operator"},
     ...
  }
}
   //Other sources ...

\end{Verbatim}
\caption{Structure of the JSON ``query" object}
%\caption{JSON constructed to send the structure of a SPARQL query for an advanced query with logical operators}
\label{fig:JSON_advquery}
\end{figure}

@@

When the results are returned, each result set is structured as the JSON object given in Figure \ref{fig:JSON_runquery}. This object identifies the endpoint where the query was executed, the URI of the endpoint, the query variables,  how many results are returned and the result set.

When the user runs an advanced query, the payload to be sent to the proxy server to perform the SPARQL query construction and query execution orchestration is bit more complex than in the ``Query All" case. The system must specify what data sources are to be queried, what properties are queried, the values for each of the selected properties, and how these property-value pairs relate to the original user-specified query term. This information is passed as a JSON object, whose structure is given in Figure \ref{fig:JSON_advquery}, to the proxy server.

@@

We believe that the unstructured keyword search is likely to encourage users without previous experience running queries on SPARQL endpoints to explore the additional features of our system that allow more flexible and advanced queries.  For example, physicians may be more willing to invest time into learning how to find better search results once they see that the system can easily provide them with some initial useful query results using the basic keyword search option. However, the execution of queries using QueryMed will be very different from a user issuing a web search on a search engine such as Google because it exploits the underlying structure of the data exposed by the individual data endpoints.

@@

The SMART query tool is a web-based application designed to allow biologists to run SPARQL queries over multiple endpoints.  Queries to the SMART system are constructed using a descriptive logic query written in the natural-language like Manchester OWL syntax \cite{Battista}.  However, it cannot be expected that the end users know how  to use the Manchester OWL syntax to write queries, whereas in QueryMed, users can construct queries intuitively by giving keywords for user selected properties.

@@

Table \ref{comparison} compares selected features of the QueryMed system with other related systems. The QueryMed system was unique among the systems that we found in that it allows endpoints to be dynamically added by the user, and provides a hybrid interface that enables the users to both query and browse data.  Other features of the QueryMed system that distinguish it from similar systems include the ability both to perform keyword queries and to construct more advanced queries taking advantage of the structure of the data.  This feature increases the ease of use of our system relative to other similar systems.  Furthermore, the Javascript based user interface of the QueryMed system, implemented using the JQuery library, makes our user interface particularly attractive, easy to interact with, and capable of handling a variety of user input events.  Another unique feature of our system is the property-based advanced query interface. This enables users to take advantage of the structure of the underlying ontologies used to represent the data without prior knowledge of the ontology structures.

>>>>

In the following sections we explain the system functionality by first giving an overview of the QueryMed system architecture and the design decisions behind crucial components of the system.

@@

The QueryMed user interface is designed to be intuitive for the end user, yet flexible enough to permit a broad range of interesting queries.  The basic query interface allows the user to run simple queries, and is designed for maximal ease of use.  The advanced search capabilities allow the user to easily construct complex logical queries that take advantage of the underlying structure of the biomedical ontologies.  The user interface also allows the user to iteratively refine queries, and displays the query results, which have been retrieved from multiple SPARQL endpoints.  The main components of the user interface are as follows:

@@

\item \textbf{Basic Query Interface:}  In order to make QueryMed easy to use, but also provide a flexible system that is capable of performing a broad variety of biomedically relevant queries, we provide an uncluttered interface with only a  blank search box and two buttons as shown in Figure \ref{fig:initial_ui}. The ``Query All" button will perform a keyword-based query over pre-selected data sources. The advanced search option ``Refine Query" allows the user to view additional query options that are handled in the Advanced Query Interface.

@@

For example, after searching for "coronary artery disease" in the basic query interface, the physician will see displayed in the Result Browser a list of disease names in the Diseasome database and drugs in DailyMed and Drugbank that relate to coronary artery 
disease. She can then filter the results using additional search terms. For example, she knows that 
the route of administration of the drug that she is interested in is injection, so she filters the drug query results on the route of administration field using the query term “injection.”   The additional filtering capabilities of the Result Browser allows her  to either browse a large number of query results, or refine her search to view only the most relevant query results.

\item \textbf{Advanced Query Interface:} The Advanced Query Interface allows the user to add new data sources, and to construct a complex SPARQL query that takes advantage of the structure of the underlying biomedical ontologies.  Thus, the end user can construct a targeted query without previous knowledge of the SPARQL language or of the structure of the relevant biomedical ontologies.

When the ``Refine Query" in the Basic Query Interface is invoked, the user is provided with the default list of data sources in the system as shown in Figure \ref{fig:select}.  The user may select from this list relevant or trusted data sources that they wish to query.  The user can also dynamically add additional data sources as shown in Figure \ref{fig:add_source}, allowing users to query endpoints of interest that are not included in the default list.

A physician who is interested in using the QueryMed system to find relevant clinical trials for 
her patient can use the Advanced Query Interface to add additional relevant datasources .  For example, she might want to use the clinical trial database LinkedCt, which is 
not in the default set of endpoints.  So, she selects the “Refine 
Query” option to select the endpoints to search, and then the “Add” option 
to include an additional endpoint. After entering the name 
and URL of the LinkedCT endpoint, she can able to search 
for clinical trials for which her patient may be eligible.

@@

When a data source is selected, the properties list is automatically populated with all distinct properties available at the selected endpoint. If the user wants to restrict her query based on one or more properties available, she can specify values for the desired properties.  Because the QueryMed system automatically displays a list of properties to the user, users can improve their queries using the underlying structure of the data to obtain more relevant query results. 
%The properties list is generated by running a SPARQL query of the form given in Figure \ref{fig:SPARQL_props} at the specified data source.

@@

The user can choose to perform exact queries, or filter results on specific keywords. If the user does not know the specific value for a property, she can instead specify a keywords using the “FILTER” option. She also can choose logical operators to connect the various parts of their query. When “AND” is used it will be appended as a basic triple pattern to the SPARQL query, i.e. as a conjunction. When “OR” is used, the specified graph pattern is made to disjunct with the rest of the query with the SPARQL UNION operator. If no logical operator is specified, “AND” will be used by default.  The advanced query feature is capable of dynamically constructing complex SPARQL queries, such as the query shown in Figure \ref{fig:SPARQL_adv}.

For example, the physician might want to further refine her search for cardiovascular diseases. She 
uses the advanced search option to construct a boolean query 
that takes advantage of the underlying structure of the RDF 
data in the database. She searches Diseasome for a list of 
diseases whose class is “Cardiovascular” or for which the associated gene is “ABCA1.” Using the QueryMed advanced 
search interface, the complex SPARQL query corresponding 
to her question is automatically constructed (Figure \ref{fig:SPARQL_adv}), and she can 
view the query results conveniently displayed in the Results Browser.

@@

\item \textbf{Dynamic Facet Creator/Filter:} The Dynamic Facet Creator/Filter allows the user to select a  set of  data sources, load the properties available at these sources, and dynamically construct a complex SPARQL query connected by logical operators to take advantage of the structure of the data at each endpoint (Figure  \ref{fig:facet_creator}). The interface allows displaying of many property lists from different sources simultaneously (grouped by source). Therefore, only information relevant to the endpoint the user is currently examining will be visible at any given time.

@@

The proxy server acts as an intermediary between the user interface and the remote SPARQL endpoints.  Its functionality is twofold:

\begin{enumerate}
\item Execute the SPARQL queries at the relevant remote SPARQL endpoints and consolidate the results to be presented in the user interface.
\item Cache the results of the current query so that refinements of the query will have reduced network and query execution latency.
\end{enumerate}

@@

\item \textbf{Translator:} The Translator is responsible for translating the user query into valid SPARQL syntax.  The Translator obtains the parameters to construct the query from the input the user specifies in the Query Interface, and dynamically constructs a SPARQL query based on the user input. The Translator relies on two services:

@@

\item \textbf{Run Query:} The purpose of this service is to execute SPARQL queries generated by the Translator, and return  the query results as a JSON object.

%\begin{figure}
%\begin{Verbatim}[frame=single]
%{
%   "bindings":
%  [
%        {"source" : "source_label",
%         "uri" : "http://source.com/sparql.",
 %    	"vars" : ["variable1","variable2"],
  %   	"count":2,
   %  	"results" : 
   %  	 [
  %   	    {"variable1": "Value 1",
%	      "variable1": "Value 2"},
 %    	    {"variable1": "Another Value 1",
  %             "variable1": "Another Value 2"},
 %  	    ...	  		
  %   	 ]
  %   	}
  %   	,
%	{ //Another Source and Result Description
%	...
%	},
%	...
 %    ]
% }
%\end{Verbatim}
%\caption{Structure of the JSON ``result" object}
%\label{fig:JSON_runquery}
%\end{figure}

\item \textbf{Advanced Queries:} This service takes as input list of sources, properties, query terms, and logical operators, which are passed to the proxy server as a JSON object.

%\begin{figure}
%\begin{Verbatim}[frame=single]
%{
% "http://some-source/sparql": 
%  {
%     "property1": {"value1":  "logical operator"},
%     "property2": {"value 2": "logical operator"},
%     "property2": {"value 3": "logical operator"},
%     ...
%  }
%}
%   //Other sources ...

%\end{Verbatim}
%\caption{Structure of the JSON ``query" object}
%\caption{JSON constructed to send the structure of a SPARQL query for an advanced query with logical operators}
%\label{fig:JSON_advquery}
%\end{figure}

@@

When the results are returned, each result set is structured as a JSON object. This object identifies the endpoint where the query was executed, the URI of the endpoint, the query variables,  how many results are returned and the result set.

When the user runs an advanced query, the proxy server takes as input the data sources to be queried, the properties to be queried, the values for each of the selected properties, and the relationship between the property-value pairs and the original user-specified query term. This information is passed to the proxy server as a JSON object.

@@

The basic query interface is designed to be as simple as possible, so that users who have little previous experience running queries on SPARQL endpoints can easily and  explore rapidly find query results.  The advanced query functionality of our system allows users the flexibility to construct complex SPARQL queries.  The QueryMed system allows the user to construct queries that take advantage of the underlying structure of the data exposed by the individual data endpoints, without requiring previous knowledge of this structure.

@@

The SMART query tool is a web-based application designed to allow biologists to run SPARQL queries over multiple endpoints.  Queries to the SMART system are constructed using a descriptive logic query written in the natural-language like Manchester OWL syntax \cite{Battista}.  However, it cannot be expected that the end users know how  to use the Manchester OWL syntax to write queries, whereas in QueryMed, users can construct queries intuitively by specifying keywords for user selected properties.

@@

Table \ref{comparison} compares selected features of the QueryMed system with other related systems. The QueryMed system was unique among the systems that we found in that it allows endpoints to be dynamically added by the user, and provides a hybrid interface that enables the users to both query and browse data.  Other features of the QueryMed system that distinguish it from similar systems include the ability both to perform keyword queries and to construct more advanced queries taking advantage of the structure of the data.  This feature increases the ease of use of our system relative to other similar systems.  Furthermore, the Javascript based user interface of the QueryMed system, implemented using the JQuery library, makes our user interface particularly attractive, easy to interact with, and capable of handling a variety of user input events.  Another unique feature of our system is the property-based advanced query interface. This interface enables users to take advantage of the structure of the underlying ontologies used to represent the data without prior knowledge of the ontology structures.
